Tools for Identifying Partisan Gerrymandering (PA)

### 2019-06-15 Latex/_maintext.tex
Update _maintext.tex

    To facilatate comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole. We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 

@@

    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government” criteria further. Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole. 

@@

        \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this ``fracking”, to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find fracking a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool. Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the U.S. District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])).}

@@

    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally” tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters (see below). Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised” pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander” -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".

@@

    Thus, the discrepancy between Republican vote share and Republican seat share was huge, and yet the seat share appeared to be immutable.

@@

    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below). Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016. 

@@

    We show in Table \ref{tab:congsum} the results for the 2011 plan found unconstitutional. Results of the four measures for the 2011 plan are based on using the average of actual congressional results for the 2012-16 period under that plan. \underline{Positive} values favor Republicans (add to their seat share), while \underline{negative} values favor Democrats (subtract from Republican seat share).   

@@

	To estimate these metrics for plans that have no electoral data associated with them, we add some additional complexity building on the ideas of \citet{GelmanKing1994_unifiedAJPS} on simulating potential outcomes in a probabilistic fashion. We adapted the code from \citet{Kastellec_et_al_2008_PS}, which parallels that used by \texttt{JudgeIt}. It models inter-election shifts and adjusts for differences over time in vote shares among and between districts in a probabilistic fashion, using a simulation with 1,000 repetitions (see more below). We shift the districts to approximate each party recieving exactly 50\% of the state-wide vote. The same procedure is applied to the 2011 enacted map so direct comparisons between the plans can be made.
    	\footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that, with only 18 districts, too much weight should not be placed on the exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.} 
    We emphasize that a partisan gerrymandering claim based on political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander"), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable. With only 18 districts, sensitivity to chance effects is particularly important. 
    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, is found in the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10\% margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.

@@

    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and that is most readily done with data on partisan outcomes for state-wide elections projected into the new districts. For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps. After extensive investigation of alternative prediction models and their accuracy in reproducing the partisan outcomes of the 2012, 2014, and 2016 congressional elections at the district level (data omitted for space reasons), we have opted for a model that takes the average of five state-wide 2016 elections that reflect a balance of Republican and Democratic success: President, U.S. Senate, PA Attorney General, PA Treasurer, and PA Auditor. Republicans won the statewide vote in two of them (President, U.S. Senate), while the Democrats won the statewide vote in three of them (Auditor, Treasurer, and Attorney General). The average vote for the Republicans across these five elections was 49.1\%.

@@

        \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level. Both ways result in the same district partisanship breakdowns in all plans.}

@@

    \footnote{We report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation among districts. This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf. \cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}. In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes. In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}

@@

    We aggregate voting district level (precincts) data to the congressional district level.

@@

    Additionally, each district's outcome is identical to the 2016 congressional district results, when we apply a uniform swing to match the results from the state-wide elections to the results from the congressional elections. We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a ``normal vote" baseline that must be ``corrected" for incumbency effects if we wish to actually predict outcomes.

@@

     We wish to do a further robustness check on the plausibility of using our five-election composite to calculate the various metrics designed to measure potential gerrymandering. One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map. What we are doing is comparing a hypothetical ``normal vote” \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections. The congressional results for 2012-2016 and an average of the three, reported in Table \ref{tab:congsum}, are 0.11 for partisan bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for \textit{declination}. The comparable results from Table \ref{tab:gerry} are 0.09, 0.24, 0.05, and 0.39. It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 49.1\% while the other is based on projected election results set to yield a 50-50 two-party vote distribution. If instead we calculate these four metrics using the composite five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24, 0.05, 0.36.

@@

%    Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map. Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of 7R-11D.

@@

		\footnote{Six Republicans and five Democrats ran for re-election after the new map was put in place. No incumbent lost. There were two Republican incumbents who won by less than 2 percentage points, along with one who won by less than 3 percentage points. There was just one seat won by Democrats in 2018 with a margin of 5 percentage points or less. Republicans therefore had four incumbents running and winning in competitive districts compared to just two for Democrats.}

@@

@@

    The findings from Table \ref{tab:gerry} are very clear; Regardless of which metric we examine, the 2011 map is the most gerrymandered. Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. All four measures of potential gerrymandering are statistically significant for the 2011 map. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. Indeed, none of the four measures of potential gerrymandering is statistically significant for the court-ordered map. And when we turn to the Republican Joint Legislative plan we find that all four measures are again statistically significant.  

@@

    In contrast, we may wish to use the term ``fair" as plans that seek to compensate for the difference in level of partisan concentration. Fair plans do not seek to achieve proportional representation. Rather, they seek to achieve partisan symmetry. A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 55\% of the vote, party L can similarly win 70\% of the districts if it had received 55\% of the vote.

@@

    One clear finding of our analysis of the court-ordered map is that it was drawn as a ``neutral map", though not necessarily a fair map in the sense we are using the term above. In order to create a truly ``fair map", in our view, it would have been necessary to break up Philadelphia County in more than three pieces so as to avoid the ``natural" packing of Democratic voters.

@@

    While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op. pp.116-117), the court took what we regard as a rather unusual step. It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state. As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.

@@

    In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland. On the other hand, we should also note that a ``''free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future. As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S. Constitution does \citep{Douglas2014_RightToVote}. In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and these clauses could, in principle, be used in exactly the same way as the ``free and equal” clause. 

@@

    Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay. If so, there remains the issue of whether state courts have the legal tools to deal with these ``\textit{stealth gerrymander}”.

@@

        \footnote{Commenting on the potential for the North Carolina district court to strike down their maps as unconstitutional (which ultimately they did), David Daley, author of ``Ratf**ked: Why Your Vote Doesn't Count" \cite{Daley2016_ratf**ked}, stated about a possible remedial plan by the map's original consultant that, ``[Thomas B. Hofeller] was really good at maps that look more normal, that break up fewer towns and counties but are just as partisan and just as advantageous for Republicans. It wouldn’t surprise me at all if on a hard drive somewhere in Raleigh Tom Hofeller has another set of gifts for legislators.” Originally cited in \citet{Barnett2018_NC_gerry}. As we have pondered in this essay about Pennsylvania, the North Carolina court would have had to grapple with a plan which, according to traditional criteria, doesn't meet the standards of a gerrymander, but when looking at the partisan outcomes, clearly does. That a plan which was created which accomplished the partisan goals of the redistricters but also shielded itself from agitators by hiding the obvious indicators of gerrymandering was not enacted over one that is vulnerable to litigation is beyond the current authors' imagination.}

@@

% 
% 
% 
% 
% 
% % ================================================================= 
% % -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % -----------------------------------------------------------------
%         \footnote{Because of its high number of competitive seats, a point prediction such as 9R-9D is somewhat misleading for the court-ordered map. Using the five-election model as our predictor, there are fewer safe Democratic seats (5) than safe Republican seats (7) in the Court plan, using 45\%-55\% definition of a competitive seat. This is one fewer safe seat for Republicans and one additional safe Democratic seat than is found in the 2011 plan, but the Court plan also has five districts balanced on a knife edge, with a projected winner’s margin under two percentage points, making it highly responsive to shifting voter sentiment. This is realized in our simulations where the mean seat share, though near 9R-9D, is either lower or higher depending on where the state-wide vote is centered.}
% % ----------------------------------------------------------------- 
% % -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% % =================================================================

% 	Past congressional results alone often are departures from the underlying district partisanship several reasons. The vote share of an incumbent has historically been greater than it is for a non-incumbent or two non-incumbents running for an open seat.

% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}. Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================

>>>>

    To facilitate comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole. We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 

@@

    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government'' criteria further. Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole. 

@@

        \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this ``fracking'', to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find fracking a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool. Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the U.S. District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])).}

@@

    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally'' tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters (see below). Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised'' pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander'' -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".

@@

    Thus, the discrepancy between Republican vote share and Republican seat share was huge, and yet the seat share appeared to be immutable (see Table \ref{tab:congsum}).

@@

    Table \ref{tab:congsum} shows other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below). Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016. 

@@

    \underline{Positive} values favor Republicans , while \underline{negative} values favor Democrats.   

@@

@@

    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and one such way is with data on partisan outcomes for state-wide elections projected into the new districts. For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps. After extensive investigation of alternative prediction models and their accuracy in reproducing the partisan outcomes of the 2012, 2014, and 2016 congressional elections at the district level (data omitted for space reasons), we have opted for a model that takes the average of five state-wide 2016 elections that reflect a balance of Republican and Democratic success: President, U.S. Senate, PA Attorney General, PA Treasurer, and PA Auditor. Republicans won the statewide vote in two of them (President, U.S. Senate), while the Democrats won the statewide vote in three of them (Auditor, Treasurer, and Attorney General). The average vote for the Republicans across these five elections was 49.1\%.

@@

        \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level (see discussion on alternative ways to tabulate vote-shares in \citet{Grofman_et_al_1997__IntedgratedPerspective_ES}. Both ways result in the same district partisanship outcomes in all plans.}

@@

    \footnote{We report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote share, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation among districts. This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf. \cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}. In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes. In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades \citep[cf.][]{Jacobson2015}.}

@@

    We aggregate voting district level (precincts) data for these five statewide elections to the congressional district level.

@@

    Additionally, each district's outcome is identical to the 2016 congressional district results, when we apply a uniform swing to match the results from the state-wide elections to the results from the congressional elections. 

@@

    We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a ``normal vote" baseline that must be ``corrected" for incumbency effects if we wish to actually predict outcomes.
     We wish to do a further robustness check on the plausibility of using our five-election composite to calculate the various metrics designed to measure potential gerrymandering. One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map. What we are doing is comparing a hypothetical ``normal vote'' \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections. The congressional results for 2012-2016 and an average of the three, reported in Table \ref{tab:congsum}, are 0.11 for partisan bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for \textit{declination}. The comparable results from Table \ref{tab:gerry} (see below) are 0.09, 0.24, 0.05, and 0.39. It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 49.1\% while the other is based on projected election results set to yield a 50-50 two-party vote distribution. If instead we calculate these four metrics using the composite five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24, 0.05, 0.36.

@@

@@

        \footnote{Six Republicans and five Democrats ran for re-election after the new map was put in place. No incumbent lost. There were two Republican incumbents who won by less than 2 percentage points, along with one who won by less than 3 percentage points. There was just one seat won by Democrats in 2018 with a margin of 5 percentage points or less. Republicans therefore had four incumbents running and winning in competitive districts compared to just two for Democrats (10+ percentage point margin).}

@@

        \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that, with only 18 districts, too much weight should not be placed on the exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay. We have devoted more space to partisan bias than to the other measures of gerrymandering reported in this essay due to it's complexity than to the other more straightforward measures.} 
    We emphasize that a partisan gerrymandering claim based on statewide political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander"), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable. With only 18 districts, sensitivity to chance effects is particularly important. 
    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, is found in the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10 percentage point margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.

@@

    The findings from Table \ref{tab:gerry} are very clear; regardless of which metric we examine, the 2011 map is the most gerrymandered. Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. All four measures of potential gerrymandering are statistically significant for the 2011 map. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. Indeed, none of the four measures of potential gerrymandering is statistically significant for the court-ordered map. And when we turn to the Republican Joint Legislative plan we find that all four measures are again statistically significant.  

@@

    In contrast, we may wish to use the term ``fair" as plans that seek to compensate for the difference in level of partisan concentration. Fair plans do not seek to achieve proportional representation. Rather, they seek to achieve partisan symmetry. A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 56\% of the vote, party L can similarly win 70\% of the districts if it had received 56\% of the vote.

@@

    One clear finding of our analysis of the court-ordered map is that it was drawn as a ``neutral map", though not necessarily a ``fair map'' in the sense we are using the term above. In order to create a truly ``fair map", in our view, it would have been necessary to break up Philadelphia County in more than three pieces so as to avoid the natural packing of Democratic voters.

@@

    While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op. pp.116-117), the court took what we regard as a rather unusual step. It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state. As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal'': Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.

@@

    In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland. On the other hand, we should also note that a ``free and equal'' elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future. As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S. Constitution does \citep{Douglas2014_RightToVote}. In addition to the thirteen states that require elections to be ``free and equal'', an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open'', and these clauses could, in principle, be used in exactly the same way as the ``free and equal'' clause. 

@@

    Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay. If so, there remains the issue of whether state courts have the legal tools to deal with these ``\textit{stealth gerrymander}''.

@@

        \footnote{Commenting on the potential for the North Carolina district court to strike down their maps as unconstitutional (which ultimately they did), David Daley, author of ``Ratf**ked: Why Your Vote Doesn't Count" \cite{Daley2016_ratf**ked}, stated about a possible remedial plan by the map's original consultant that, ``[Thomas B. Hofeller] was really good at maps that look more normal, that break up fewer towns and counties but are just as partisan and just as advantageous for Republicans. It wouldn’t surprise me at all if on a hard drive somewhere in Raleigh Tom Hofeller has another set of gifts for legislators.'' Originally cited in \citet{Barnett2018_NC_gerry}. As we have pondered in this essay about Pennsylvania, the North Carolina court would have had to grapple with a plan which, according to traditional criteria, doesn't meet the standards of a gerrymander, but when looking at the partisan outcomes, clearly does. That a plan which was created which accomplished the partisan goals of the redistricters but also shielded itself from agitators by hiding the obvious indicators of gerrymandering was not enacted over one that is vulnerable to litigation is beyond the current authors' imagination.}

@@

% 
% 
% 
% 
% 
%   Past congressional results alone often are departures from the underlying district partisanship several reasons. The vote share of an incumbent has historically been greater than it is for a non-incumbent or two non-incumbents running for an open seat.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
%       \footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}. Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
